old equations

### 2022-04-27 reply-to-review1/old-eqns.Rmd
updated bib file to match Rmd citations

---
title: "Mendelian randomization with genome-wide association studies data"
author: "Frederick J. Boehm & Xiang Zhou"
date: "`r lubridate::now()`"
bibliography: exported.bib
csl: elsevier-with-titles.csl
link-citations: yes
linkcolor: blue
always_allow_html: true
output: 
  - bookdown::pdf_document2
  - bookdown::html_document2
  - bookdown::word_document2
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
library(magrittr)
```

# Abstract

Complex trait genetics studies in humans have yielded thousands of new associations for  
diverse traits. The next step in understanding biology of health and disease is to
identify causal relationships among complex traits. Two-sample Mendelian randomization studies 
in genome-wide association studies can uncover causal relationships 
between complex traits. By leveraging shared SNP genotypes as instrumental variables, or proxies, for 
the exposure complex trait, investigators can tease out causal effects from observational data, provided
that necessary assumptions are satisfied. We discuss below the growth of genome-wide association studies 
in parallel with development of Mendelian randomization methods. We argue that the recent availability of GWAS summary statistics for diverse complex traits has motivated new Mendelian randomization methods and that this area continues to offer opportunities for additional methods developments.

# Introduction

Great interest has developed in inferring causal relationships between complex 
traits, *i.e.*, traits that seemingly are not inherited in a Mendelian fashion, in observational human genetics studies. Discovery of such relationships is crucial to 
enhancing our understanding of the biology of health and disease. Mendelian randomization (MR) studies are a form of instrumental variable analysis in which
genetic markers, typically single nucleotide polymorphisms (SNPs), serve as instruments, or proxies, for 
inferring causal effects of an exposure variable on an outcome variable [@wright1928tariff;@wright1921correlation;@wright1923theory;@stock2003retrospectives;@katan1986apolipoprotein;@katan2004commentary;@didelez2007mendelian].

Genome-wide association studies (GWAS) present unique opportunities for 
discovery of causal relationships via MR [@swerdlow_selecting_2016]. A GWAS
interrogates millions of single nucleotide polymorphisms (SNPs) to infer which affect the trait [@the_wellcome_trust_case_control_consortium_genome-wide_2007;@klein_complement_2005].
In the nearly 17 years since publication of early studies, researchers have reported 
thousands of novel SNP-trait associations from GWAS [@klein_complement_2005;@welter_nhgri_2014;@visscher_10_2017;@visscher_five_2012]. Ever larger sample sizes in GWAS have enhanced statistical power to detect associations and have refined our understanding of 
human health and disease.

```{r, article-counts, message = FALSE, fig.cap= "Upward trend in article counts by year for keyword searches: 1. mendelian randomization and 2. genome-wide association study."}
library(magrittr)
dat <- readxl::read_xlsx(path = "data/publication-counts.xlsx", sheet = "counts")
dat %>%
  tidyr::pivot_longer(!year) %>%
  dplyr::filter(year < 2022) %>%
  ggplot2::ggplot(ggplot2::aes(x = year, y = value, fill = name)) + 
  ggplot2::geom_bar(stat = "identity", 
                    position=ggplot2::position_dodge()) + 
  ggplot2::scale_fill_manual(labels = c("GWAS", "MR"), values=c('#999999','#E69F00')) + 
  ggplot2::ylab("Annual publication count") + 
  ggplot2::xlab("Year") + 
  #ggplot2::ggtitle("Published journal articles per year for Google Scholar key word searches: 1. Mendelian randomization and 2. Genome-wide association study") + 
  ggplot2::guides(fill = ggplot2::guide_legend("Search terms")) + 
  ggplot2::theme(text = ggplot2::element_text(size = 20))

ggplot2::ggsave("figures/mr-gwas-counts.png")

A number of excellent MR reviews have appeared in the literature [@burgess_review_2017;@richmond_concepts_2022;@sanderson_mendelian_2022]. Our review doesn't 
aim to replace them, but seeks to complement their findings. 
We present recent methods developments for an audience of statisticians and epidemiologists. We discuss in detail methods advances before briefly summarizing 
applications and directions for future research. The value of MR analysis with 
GWAS data will continue to grow in parallel with statistical methods advances.

Methods for and uses of MR have appeared at a rapid and accelerating pace since the 
publication of the early GWAS (Figure \@ref(fig:article-counts)). Many studies use the "two-sample" MR 
design, in which one cohort of subjects has measurements for the exposure, while a second cohort has 
measurements for a second trait [@pierce_efficient_2013]. Importantly, the SNP instruments are measured in 
both sets of subjects.

Recent methods developments have enabled execution of two-sample MR studies with only 
GWAS summary statistics [@burgess2015using]. These strategies are especially pertinent now that there is 
an abundance of freely available GWAS summary data [@buniello2019nhgri]. The 
ability to use published data has contributed to a rapid increase in the number of
two-sample MR methods for GWAS summary statistics (Figure \@ref(fig:article-counts)).

# Mendelian Randomization with one Instrument

We begin our discussion of Mendelian randomization in GWAS data by considering the simplest case, where 
there is a single genetic marker and a single outcome variable. Mendelian randomization with one 
instrument uses a single genetic marker 
as the proxy ("instrument" or "instrumental variable") for the "exposure", a complex trait, and asks 
whether variation in the genotypes at the genetic marker causes variation in the 
"outcome" variable, which is typically a second complex trait. When the genetic marker 
is a SNP on an autosome, the cohort is partitioned into three 
classes, two homozygote classes and a single heterozygote class, based on genotypes 
for that marker. When the three assumptions of mendelian randomization are satisfied, 
estimation of the causal effect of the exposure on the outcome is possible.

The outcome variable can be binary, a count, a time to event, or a continuous variable. For brevity and simplicity, 
we focus on continuous outcomes before considering other classes of outcome variables in a later section.

## Assumptions of Mendelian Randomization

Three assumptions are central to MR. Satisfying them enables inference of 
causal effects in an observational MR study. In practice, careful evaluation of the assumptions is needed. 
Because assumptions 2 and 3, which involve absence of certain types of confounding, can't be verified, 
practitioners must be cautious when planning MR analyses. Sensitivity analysis is highly recommended as a 
way to assess robustness of estimates in the presence of possible assumption violations [@vanderweele_methodological_2014].

The first assumption states that the exposure is associated with the outcome. The second requires 
independence, conditional on the exposure and all measured and unmeasured confounders, between the 
genetic marker and the outcome. The last of the three assumptions states that the genetic marker is 
independent of all (measured and unmeasured) confounders of the relationship between exposure and 
outcome. We next discuss these in more detail and consider their evaluation in MR studies [@vanderweele_methodological_2014].

If the first assumption, about the association between the genetic marker and the exposure, holds, 
but this association is weak, then an amplification of biases, such as those due to violations of 
assumptions 2 and 3, may result [@vanderweele_methodological_2014]. @burgess_avoiding_2011, in analyses 
of simulated and real data, demonstrated that bias in causal estimates increases with decreases in 
the strength of association between the SNP and exposure. We later discuss, in the context of 
methods for multiple SNP instruments, how this finding, coupled with the recognition of 
the polygenicity of many complex traits, has motivated methods to work with large numbers of weak SNP instruments [@ye_debiased_2021;@chao_consistent_2005].

The second MR assumption, often termed the exclusion restriction assumption, can also be stated as 
the need for the genetic marker to affect the outcome only through the exposure. In a 
causal diagram, this means that the only path from the marker to the outcome is that 
containing the exposure.

The third assumption ensures that the genetic marker is
independent of all confounders of the exposure-outcome relationship. Like the exclusion restriction
assumption, we can't verify this assumption with observational data, so analysts need to 
use tools like sensitivity analysis to ensure validity of findings.

When the three MR assumptions hold, one may use instrumental variable statistical methods, 
with the genetic marker data, the exposure data, and the outcome data, to estimate and to test the 
causal effect of the exposure on the outcome [@didelez_mendelian_2007;@lawlor_mendelian_2008;@palmer_adjusting_2008].

To understand the origin of the term "Mendelian randomization", it is helpful to 
recall Mendel's inheritance laws [@mendel1865versuche]. Under the assumption that alleles segregate 
randomly from parent to offspring, the offspring genotypes are unlikely to be 
associated with confounders of the exposure-outcome relationship. Additionally,
reverse causation is unanticipated since germ-line genotypes are 
fixed at conception and, thus, precede realization of other observed variables. Reverse causation 
occurs when the outcome at an early time point has a causal effect on the exposure variable. While 
reverse causation is often assumed to not occur in MR study designs, @burgess_are_2021 points out 
several scenarios where this assumption may not hold [@davey_smith_mendelian_2003].

## Statistical models and methods for MR with one instrument

Approaches to MR with one SNP instrument can be classified into four categories: ratio of coefficients method, two-stage methods, 
likelihood-based methods, and semiparametric methods [@burgess_review_2017]. We discuss each of these before turning to methods 
that leverage multiple SNP instruments.

### Ratio of coefficients method

The ratio of coefficients method, also known as the Wald method, estimates the 
causal effect of the exposure X on the outcome Y by using a single SNP instrument [@wald_fitting_1940]. 
For a continuous outcome, the causal effect estimator, $\hat\gamma_{Wald}$ is defined in Equation \@ref(eq:wald).

@@

In Equation \@ref(eq:wald), $\hat \beta_{YZ}$ and $\hat \beta_{XZ}$ are the slope 
estimates from the regressions of the outcome and exposure, respectively, on the 
SNP instrument. Since the Wald method requires only the regression coefficients, it can be used with 
summary data. However, the Wald method doesn't accommodate multiple SNP instruments, which limits its 
direct use in the GWAS setting.

@@

A two-stage statistical model involves two regression models [@burgess_review_2017;@sawa_exact_1969]. For a continuous 
outcome, one may perform "two-stage least squares", which involves two linear regressions. 
First, the exposure variable is regressed on the instrument (*i.e.*, the SNP genotypes) (Equation \@ref(eq:lm1)). We denote each subject's exposure variable value as $x_i$ in Equation \@ref(eq:lm1), while $\alpha_0$ denotes an intercept term and $\alpha_1$ is the slope. $z_i$ denotes the SNP instrument genotype for subject $i$, while $\epsilon_i$, the error term, is assumed to be independent among subjects and normally distributed with a shared common variance and mean zero.

@@

# References

>>>>

---
title: "old equations"
author: "Frederick Boehm"
date: "4/27/2022"
output: html_document
---

For a continuous outcome, the causal effect estimator, $\hat\gamma_{Wald}$ is defined in Equation \@ref(eq:wald).

@@

In Equation \@ref(eq:wald), $\hat \beta_{YZ}$ and $\hat \beta_{XZ}$ are the slope 
estimates from the regressions of the outcome and exposure, respectively, on the 
SNP instrument.

@@

First, the exposure variable is regressed on the instrument (*i.e.*, the SNP genotypes) (Equation \@ref(eq:lm1)). We denote each subject's exposure variable value as $x_i$ in Equation \@ref(eq:lm1), while $\alpha_0$ denotes an intercept term and $\alpha_1$ is the slope. $z_i$ denotes the SNP instrument genotype for subject $i$, while $\epsilon_i$, the error term, is assumed to be independent among subjects and normally distributed with a shared common variance and mean zero.

@@
